---
title: "hw02-ntjjmak"
author: "Nicole Mak"
date: "24/09/2018"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Let's load the data.

```{r loading}
library(gapminder)
library(tidyverse)
```

**What kind of data is this? Print it out.**

```{r characterising the data}
typeof(gapminder)
class(gapminder)
gapminder
```



```{r basic exploration}
str(gapminder)
ncol(gapminder)
nrow(gapminder)
```


**Let's examine life expectancy and gdp Per capita. What are the values, ranges, distribution to be expected by continent? Here are some summary tables:**

```{r individual variable exploration}
gapminder %>% 
  select(continent)%>% 
  summary()
gapminder %>% 
  select(lifeExp) %>% 
  summary()
```

**Values, ranges, distribution can also be explored visually using graphs. Let's also use a figure to examine the distribution and frequency of values for the variable "Life Expectancy".**

```{r basic univariable histogram}
gapminder %>% 
  ggplot(aes(lifeExp)) +
  geom_histogram(binwidth = 1)
```

**Let's try a density plot.**

```{r basic univariable density plot}
gapminder %>% 
  ggplot(aes(lifeExp)) +
  geom_density(kernel = "gaussian")
  
```

**Let's do a further exploration of life Expectancy and gdpPercap by continent.**

First, we can explore by boxplot.

```{r experiment with 2 variables}
gapminder %>% 
  ggplot(aes(continent, lifeExp)) + 
  geom_boxplot()

gapminder %>% 
  ggplot(aes(continent, gdpPercap)) + 
  geom_boxplot()
```

**The boxplot exploring gdpPercap by continent is a bit difficult to interpret. Let us transform it to a log scale.**

```{r log transformation trial}
gapminder %>% 
  ggplot(aes(continent, log(gdpPercap))) + 
  geom_boxplot()
```


**To get an idea of which values are typical, we can alternatively use a scatter plot so that all data points are seen. Transparency is added to give an idea of which values are most frequent in the data set.**

```{r experimenting with transparency}
gapminder %>% 
  ggplot(aes(continent, log(gdpPercap))) + 
  geom_point(alpha = 0.05)
```

**Here is some practice filtering to single countries, Canada and China.Let's see whether these countries see a rise in life expectancy over the years.** 

```{r applying filtering and piping trial 1}
gapminder %>% 
  filter(country == "Canada") %>% 
  ggplot(aes(year, lifeExp)) +
  geom_point()

gapminder %>% 
  filter(country == "China") %>% 
  ggplot(aes(year, lifeExp)) +
  geom_point()
```

**Here is some more practice piping.**

```{r building more piping and incorporating log transformation}
gapminder %>% 
  filter(continent == "Oceania") %>% 
  filter(year > "1990") %>% 
  ggplot(aes(country, log(gdpPercap))) +
  geom_point()

```

```{r again experimenting with piping}
gapminder %>% 
  select(continent, lifeExp, gdpPercap) %>% 
  filter(lifeExp >= "75") %>% 
  ggplot(aes(continent, log(gdpPercap))) +
  geom_violin(fill = "black")
```


**Let's play with colour.**

```{r more visualisation but with colour}
gapminder %>% 
  filter(year > "1970") %>%
  ggplot(aes(continent, mean(pop), fill = continent)) +
  geom_col()
```

**Let's try flipping it around.**

```{r flipping x and y}
gapminder %>% 
  filter(year > "1970") %>%
  ggplot(aes(continent, mean(pop), fill = continent)) +
  geom_col() + 
  coord_flip()
```


**Lastly, let's experiment a couple more `dplyr` functions.**

```{r just more dplyr trials}
select(gapminder, -continent, - gdpPercap) %>% 
  filter(year == "2007") %>%
  arrange(lifeExp)
```


## All done.
