---
title: "hw02-ntjjmak"
author: "Nicole Mak"
date: "24/09/2018"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Let's load the data.

```{r cars}
library(gapminder)
library(tidyverse)
```

What kind of data is this? Print it out.

```{r}
typeof(gapminder)
class(gapminder)
gapminder
```



```{r}
str(gapminder)
ncol(gapminder)
nrow(gapminder)
```


Let's examine life expectancy and gdp Per capita. What are the values, ranges, distribution to be expected by continent?

Here are some summary tables:

```{r}
gapminder %>% 
  select(continent)%>% 
  summary()
gapminder %>% 
  select(lifeExp) %>% 
  summary()
```

Values, ranges, distribution can also be explored visually using graphs.

Let's also use a figure to examine the distribution and frequency of values for the variable "Life Expectancy".

```{r}
gapminder %>% 
  ggplot(aes(lifeExp)) +
  geom_histogram(binwidth = 1)
```

Let's try a density plot.

```{r}
gapminder %>% 
  ggplot(aes(lifeExp)) +
  geom_density(kernel = "gaussian")
  
```

Let's do a further exploration of life Expectancy and gdpPercap by continent.

First, we can explore by boxplot.

```{r}
gapminder %>% 
  ggplot(aes(continent, lifeExp)) + 
  geom_boxplot()

gapminder %>% 
  ggplot(aes(continent, gdpPercap)) + 
  geom_boxplot()
```

The boxplot exploring gdpPercap by continent is a bit difficult to interpret. Let us transform it to a log scale.

```{r}
gapminder %>% 
  ggplot(aes(continent, log(gdpPercap))) + 
  geom_boxplot()
```


To get an idea of which values are typical, we can alternatively use a scatter plot so that all data points are seen. Transparency is added to give an idea of which values are most frequent in the data set.

```{r}
gapminder %>% 
  ggplot(aes(continent, log(gdpPercap))) + 
  geom_point(alpha = 0.05)
```

Here is some practice filtering to single countries, Canada and China. 
Let's see whether these countries see a rise in life expectancy over the years. 

```{r}
gapminder %>% 
  filter(country == "Canada") %>% 
  ggplot(aes(year, lifeExp)) +
  geom_point()

gapminder %>% 
  filter(country == "China") %>% 
  ggplot(aes(year, lifeExp)) +
  geom_point()
```

Here is some more practice piping.

```{r}
gapminder %>% 
  filter(continent == "Oceania") %>% 
  filter(year > "1990") %>% 
  ggplot(aes(country, log(gdpPercap))) +
  geom_point()

```

```{r}
gapminder %>% 
  select(continent, lifeExp, gdpPercap) %>% 
  filter(lifeExp >= "75") %>% 
  ggplot(aes(continent, log(gdpPercap))) +
  geom_violin(fill = "black")
```


Let's play with colour.

```{r}
gapminder %>% 
  filter(year > "1970") %>%
  ggplot(aes(continent, mean(pop), fill = continent)) +
  geom_col()
```

Let's try flipping it around.

```{r}
gapminder %>% 
  filter(year > "1970") %>%
  ggplot(aes(continent, mean(pop), fill = continent)) +
  geom_col() + 
  coord_flip()
```


Lastly, let's experiment a couple more `dplyr` functions.

```{r}
select(gapminder, -continent, - gdpPercap) %>% 
  filter(year == "2007") %>%
  arrange(lifeExp)
```


All done.
