---
title: "STAT545_hmk4_WilliamHall"
author: "William Hall"
date: '2018-10-04'
output: github_document
always_allow_html: yes
---


##Data Reshaping Prompts

>Activity #2

Make a tibble with one row per year and columns for life expectancy for two or more countries.

- Use knitr::kable() to make this table look pretty in your rendered homework.
- Take advantage of this new data shape to scatterplot life expectancy for one country against that of another.

The first thing we need to do is install the kable package.
```{r}
#install.packages("kableExtra")
library(knitr)
library(kableExtra)
library(gapminder)
library(dplyr)
library(readxl)
```

So let's have a look at the gapminder dataset without any processing with kable()

```{r}
(mindthegap <- gapminder[1:10, ])
```

##need to get the table displaying inline, and not in the window....

Gross. Now let's kable it.

```{r}
mindthegap %>%
  kable() %>%
  kable_styling()
```

Cute. Now let's make different coloured lines.

```{r}
kable(mindthegap) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

Okkkk. But those rows look a little big. Let's make them smaller

```{r}
kable(mindthegap) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Ok fine. But I want the table on the left.

```{r}
kable(mindthegap) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "left")
```

Just joking. I want it on the right - and I want the text beside it.

##how do I do this? Here is where I got it from: https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html

```{r}
kable(mindthegap) %>%
  kable_styling(bootstrap_options = "striped", full_width = F, position = "float_right")
```

Ok now what if I want to change font size.

```{r}
kable(mindthegap) %>%
  kable_styling(bootstrap_options = "striped", font_size = 14)
```

Cool. Now what if we want to high year 1972 and 1992.

```{r}
kable(mindthegap) %>%
  kable_styling("striped", full_width = F) %>%
  column_spec(1:6, bold = T) %>%
  row_spec(c(5,9), bold = T, color = "white", background = "#D7261E")
```

Sweet. So now lets put some headers above our columns.

```{r}
kable(mindthegap) %>%
  kable_styling(c("striped", "bordered"), full_width = F) %>%
  column_spec(1:6, bold = T) %>%
  row_spec(c(5,9), bold = T, color = "white", background = "#D7261E") %>% 
  add_header_above(c("Place" = 2, " " = 4))
```

>Activity #3

Compute some measure of life expectancy (mean? median? min? max?) for all possible combinations of continent and year. Reshape that to have one row per year and one variable for each continent. Or the other way around: one row per continent and one variable per year.

- Use knitr::kable() to make these tables look pretty in your rendered homework.
- Is there a plot that is easier to make with the data in this shape versis the usual form? If so (or you think so), try it! Reflect.

So what we are trying to do here is calculate the mean life expectancy for every combination of continent and year. Then reduce down to a table with just the year, continent, and mean. So let's use the group_by and summarise function.


```{r}
gapminder %>% 
  group_by(continent,year) %>%
  summarise(meanLife = mean(lifeExp))
```

         
##Join Prompts

>Activity #1

Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins. Examples of a second data frame you could build:

- One row per country, a country variable and one or more variables with extra info, such as language spoken, NATO membership, national animal, or capitol city.
- One row per continent, a continent variable and one or more variables with extra info, such as northern versus southern hemisphere.

So let's start with the countries, and their languages. We will begin by uploading a dataset with all the countries and their languages. Then remove all the NA's, then rename the columns (remember that the column name that you want to use to join must be identical in both datasets).

```{r}
LanguageCountry <- read_excel("LanguageCountry.xlsx")

LangCountry <- LanguageCountry %>%
                        na.omit() 
colnames(LangCountry) <- c("country", "language")
LangCountry

```

Now lets join it with the population of the gap minder dataset. Let's create the reduced gapminder dataset with only the population and country.

```{r}

reducedGap <- gapminder %>%
                group_by(country) %>%
                summarise(meanpop = mean(pop)) %>%
                mutate_if(is.factor, as.character)
```

Join the two together with full join.

```{r}
LangPopCountry <- full_join(LangCountry, reducedGap, by = "country")
LangPopCountry
```



##Reflection

Had some problems with the group_by function because I hadn't loaded the dplyr library. Spent way too much time trying to fiddle with the code. Learned that you should clear your workspace, and start new everytime so as to not make mistakes like that. 








